Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-05-30 submission/manuscript.tex
Fix email address
Make figures an explicit dependency and remove from body of manuscript
Respond to reviewers' comments

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

@@

Colonic bacterial populations are thought to have a role in the
development of colorectal cancer with some protecting against
inflammation and others exacerbating inflammation. Short-chain fatty
acids (SCFAs), including butyrate, have been shown to have
anti-inflammatory properties and are produced in large quantities by
colonic bacteria which produce SCFAs by fermenting fiber. We assessed
whether there was an association between fecal SCFA concentrations and
the presence of colonic adenomas or carcinomas in a cohort of
individuals using 16S rRNA gene and metagenomic shotgun sequence data.
We measured the fecal concentrations of acetate, propionate, and
butyrate within the cohort and found that there were no significant
associations between SCFA concentration and tumor status. When we
incorporated these concentrations into random forest classification
models trained to differentiate between people with normal colons and
those with adenomas or carcinomas, we found that they did not
significantly improve the ability of 16S rRNA gene or metagenomic gene
sequence-based models to classify individuals. Finally, we generated
random forest regression models trained to predict the concentration of
each SCFA based on 16S rRNA gene or metagenomic gene sequence data from
the same samples. These models performed poorly and were able to explain
at most 14\% of the observed variation in the SCFA concentrations. These
results support the broader epidemiological data that questions the
value of fiber consumption for reducing the risks of colorectal cancer.
Although other bacterial metabolites may serve as biomarkers to detect
adenomas or carcinomas, fecal SCFA concentrations have limited
predictive power.

@@

Colorectal cancer is the third leading cancer-related cause of death
within the United States (1). Less than 10\% of cases can be attributed
to genetic risk factors (2). This leaves a significant role for
environmental, behavioral, and dietary factors (3, 4). Colorectal cancer
is thought to be initiated by a series of mutations that accumulate as
the mutated cells begin to proliferate leading to adenomatous lesions,
which are succeeded by carcinomas (2). Throughout this progression,
there are ample opportunities for bacterial populations to have a role
as bacteria are known to cause mutations, induce inflammation, and
accelerate tumorigenesis (5--7). Additional cross sectional studies in
humans have identified microbiome-based biomarkers of disease (8). These
studies suggest that in some cases, it is the loss of bacterial
populations that produce short-chain fatty acids (SCFAs) that results in
increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence
of 16S rRNA gene sequences from organisms and the genes involved in
producing them as a biomarker of a healthy microbiota (9, 10). SCFAs
have anti-inflammatory and anti-proliferative activities (11). Direct
supplementation of SCFAs or feeding of fiber caused an overall reduction
in tumor burden in mouse models of colorectal cancer (12). These results
suggest that supplementation with fiber, which many colonic bacteria
ferment to produce SCFAs may confer beneficial effects against
colorectal cancer. Regardless, there is a lack of evidence that
increasing SCFA concentrations can protect against colorectal cancer in
humans. Case-control studies that have investigated possible
associations between SCFAs and colon tumor status have been plagued by
relatively small numbers of subjects, but have reported increased total
and relative fecal acetate levels and decreased relative fecal butyrate
concentrations in subjects with colonic lesions (13). In randomized
controlled trials fiber supplementation has been inconsistently
associated with protection against tumor formation and recurrence (14,
15). Such studies are plagued by difficulties insuring subjects took the
proper dose and using subjects with prior polyp history who may be
beyond a point of benefiting from fiber supplementation. Together, these
findings temper enthusiasm for treatments that target the production of
SCFAs or for using them as biomarkers for protection against
tumorigenesis.

\textbf{Fecal SCFA concentrations did not vary with diagnosis or
treatment.} To quantify the associations between colorectal cancer, the
microbiome, and SCFAs, we quantified the concentration of acetate,
propionate, and butyrate in feces of previously characterized
individuals with normal colons (N=172) and those with colonic adenomas
(N=198) or carcinomas (N=120) (16). We were unable to detect a
significant difference in any SCFA concentration across the diagnoses
groups (all P\textgreater{}0.15; Figure 1A). Among the individuals with
adenomas and carcinomas, a subset (N\textsubscript{adenoma}=41,
N\textsubscript{carcinoma}=26) were treated and sampled a year later
(17). None of the SCFAs exhibited a significant change with treatment
(all P\textgreater{}0.058; Figure 1B). For both the pre-treatment
cross-sectional data and the pre/post treatment data, we also failed to
detect any significant differences in the relative concentrations of any
SCFAs (P\textgreater{}0.16). Finally, we pooled the SCFA concentrations
on a total and per molecule of carbon basis and again failed to observe
any significant differences (P\textgreater{}0.077). These results
demonstrated that there were no significant associations between fecal
SCFA concentration and diagnosis or treatment.

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units (OTUs) based on 97\% similarity or into genera enabled
us to classify individuals as having adenomas or carcinomas using random
forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the SCFAs as possible features to train the
models (Figure S1). Models trained using SCFAs to classify individuals
as having adenomas or carcinomas rather than normal colons had median
areas under the receiver operator characteristic curve (AUROC) that were
significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
and P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC
values to detect the presence of adenomas or carcinomas were only 0.54
and 0.55, respectively, indicating that SCFAs had poor predictive power
on their own (Figure 2A). When we trained the models with the SCFAs
concentrations and OTU or genus-level relative abundances the AUROC
values were not significantly different from the same models trained
without the SCFA concentrations (P\textgreater{}0.15; Figure 2A). These
data demonstrate that knowledge of the SCFA profile from a subject's
fecal sample did not improve the ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} We next asked whether the fecal community structure was
predictive of fecal SCFA concentrations, regardless of a person's
diagnosis. We trained random forest regression models using 16S rRNA
gene sequence data binned into OTUs and genera to predict the
concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
between the observed SCFA concentrations and the modeled concentrations
was 0.14, which was observed when using genus data to predict butyrate
concentrations (Figure 2B). We also used a smaller dataset of shotgun
metagenomic sequencing data generated from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories and trained random forest regression models using metagenomic
sequence data to predict the concentration of the SCFAs (Figure S2).
Similar to the analysis using 16S rRNA gene sequence data, the
metagenomic data was not predictive of SCFA concentration. The largest
R\textsuperscript{2} was 0.055, which was observed when using KEGG data
to predict propionate concentrations (Figure 2B). Because of the limited
number of samples that we were able to generate metagenomic sequence
data from, we used our 16S rRNA gene sequence data to impute metagenomes
that were binned into metabolic pathways or KEGG categories using
PICRUSt (Figure S2). SCFA concentrations could not be predicted based on
the imputed metagenomic data. The largest R\textsuperscript{2} was
0.085, which was observed when using KEGG data to predict propionate
concentrations (Figure 2B). The inability to model SCFA concentrations
from microbiome data indicates that the knowledge of the abundance of
organisms and their genes was insufficient to predict SCFA
concentrations.

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not predictive of
SCFA concentrations. These results complement existing literature that
suggest that fiber consumption and the production of SCFAs are unable to
prevent the risk of developing colonic tumors. It is important to note
that our analysis was based on characterizations of SCFA and microbiome
profiles using fecal samples and that observations along the mucosa near
the site of lesions may provide a stronger association. This may be a
cautionary result to temper enthusiasm for SCFAs as a biomarker of gut
health more generally. Going forward it is critical to develop
additional hypotheses for how the microbiome and host interact to drive
tumorigenesis so that we can better understand tumorigenesis and
identify biomarkers that will allow early detection of lesions.

@@

\textbf{random forest models.} The classification models were built to
predict lesion type from microbiome information with or without SCFA
concentrations. The regression models were built to predict the SCFA
concentrations of acetate, butyrate, and propionate from microbiome
information. For classification and regression models, we pre-processed
the features by scaling them to vary between zero and one. Features with
no variance in the training set were removed from both the training and
testing sets. We randomly split the data into training and test sets so
that the training set consisted of 80\% of the full dataset while the
test set was composed of the remaining data. The training set was used
for hyperparameter selection and training the model and the test set was
used for evaluating prediction performance. For each model, the best
performing hyperparameter, mtry, was selected in an internal five-fold
cross-validation of the training set with 100 randomizations. Six values
of mtry were tested and the value that provided the largest AUROC or
R\textsuperscript{2} was selected. We trained the random forest model
using the selected mtry value and predicted the held-out test set. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to get a reliable and robust reading of model
prediction performance. We used AUROC and R\textsuperscript{2} as the
prediction performance metric for classification and regression models,
respectively. We used randomForest package implemented to the caret
package (version 4.6-14) in R statistical software (version 6.0-81) for
our models.

@@

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at \url{https://github.com/SchlossLab/Sze_SCFACRC_XXXX_2019/}.

@@

\includegraphics{figure_1.png}

>>>>

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{pschloss@umich.edu}

@@

Colonic bacterial populations are thought to have a role in the
development of colorectal cancer with some protecting against
inflammation and others exacerbating inflammation. Short-chain fatty
acids (SCFAs) have been shown to have anti-inflammatory properties and
are produced in large quantities by colonic bacteria which produce SCFAs
by fermenting fiber. We assessed whether there was an association
between fecal SCFA concentrations and the presence of colonic adenomas
or carcinomas in a cohort of individuals using 16S rRNA gene and
metagenomic shotgun sequence data. We measured the fecal concentrations
of acetate, propionate, and butyrate within the cohort and found that
there were no significant associations between SCFA concentration and
tumor status. When we incorporated these concentrations into random
forest classification models trained to differentiate between people
with normal colons and those with adenomas or carcinomas, we found that
they did not significantly improve the ability of 16S rRNA gene or
metagenomic gene sequence-based models to classify individuals. Finally,
we generated random forest regression models trained to predict the
concentration of each SCFA based on 16S rRNA gene or metagenomic gene
sequence data from the same samples. These models performed poorly and
were able to explain at most 14\% of the observed variation in the SCFA
concentrations. These results support the broader epidemiological data
that questions the value of fiber consumption for reducing the risks of
colorectal cancer. Although other bacterial metabolites may serve as
biomarkers to detect adenomas or carcinomas, fecal SCFA concentrations
have limited predictive power.

@@

Colorectal cancer is the third leading cancer-related cause of death
within the United States (1). Less than 10\% of cases can be attributed
to genetic risk factors (2). This leaves a significant role for
environmental, behavioral, and dietary factors (3, 4). Colorectal cancer
is thought to be initiated by a series of mutations that accumulate as
the mutated cells begin to proliferate leading to adenomatous lesions,
which are succeeded by carcinomas (2). Throughout this progression,
there are ample opportunities for bacterial populations to have a role
as some bacteria are known to cause mutations, induce inflammation, and
accelerate tumorigenesis (5--7). Additional cross sectional studies in
humans have identified microbiome-based biomarkers of disease (8). These
studies suggest that in some cases, it is the loss of bacterial
populations that produce short-chain fatty acids (SCFAs) that results in
increased inflammation and tumorigenesis.

Many microbiome studies use the concentrations of SCFAs and the presence
of 16S rRNA gene sequences from organisms and the genes involved in
producing them as a biomarker of a healthy microbiota (9, 10). Depending
on the concentrations, SCFAs can have proliferative activities at low
concentrations or anti-proliferative activities at higher
concentrations; they can also have anti-inflammatory activities (11).
Direct supplementation of SCFAs or feeding of fiber caused an overall
reduction in tumor burden in mouse models of colorectal cancer (12).
These results suggest that supplementation with fiber, which many
colonic bacteria ferment to produce SCFAs, may confer beneficial effects
against colorectal cancer. Regardless, there is a lack of consistent
evidence that increasing SCFA concentrations can protect against
colorectal cancer in humans. Case-control studies that have investigated
possible associations between SCFAs and colon tumor status have been
plagued by relatively small numbers of subjects, but have reported
increased total and relative fecal acetate levels and decreased relative
fecal butyrate concentrations in subjects with colonic lesions (13). In
randomized controlled trials fiber supplementation has been
inconsistently associated with protection against tumor formation and
recurrence (14, 15). Such studies are confounded by difficulties
ensuring subjects took the proper dose and using subjects with prior
polyp history who may be beyond a point of benefiting from fiber
supplementation. Together, these findings temper enthusiasm for
treatments that target the production of SCFAs or for using them as
biomarkers for protection against tumorigenesis.

\textbf{Fecal SCFA concentrations did not vary with diagnosis or
treatment.} To test for a significant association between colorectal
cancer and SCFAs, we quantified the concentration of acetate,
propionate, and butyrate in feces of previously characterized
individuals with normal colons (N=172) and those with colonic adenomas
(N=198) or carcinomas (N=120) (16). We were unable to detect a
significant difference in any SCFA concentration across the diagnoses
groups (all P\textgreater{}0.15; Figure 1A). Among the individuals with
adenomas and carcinomas, a subset (N\textsubscript{adenoma}=41,
N\textsubscript{carcinoma}=26) were treated and sampled a year later
(17). None of the individuals showed signs of recurrence and yet none of
the SCFAs exhibited a significant change with treatment (all
P\textgreater{}0.058; Figure 1B). For both the pre-treatment
cross-sectional data and the pre/post treatment data, we also failed to
detect any significant differences in the relative concentrations of any
SCFAs (P\textgreater{}0.16). Finally, we pooled the SCFA concentrations
on a total and per molecule of carbon basis and again failed to observe
any significant differences (P\textgreater{}0.077). Although some of the
P-values from our analyses were close to 0.05, the effect sizes were all
relatively small and inconsistent given the disease progression (Figure
1). These results demonstrated that there were no significant
associations between fecal SCFA concentration and diagnosis or
treatment.

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas using a random
forest model.} We previously found that binning 16S rRNA gene sequence
data into operational taxonomic units (OTUs) based on 97\% similarity or
into genera enabled us to classify individuals as having adenomas or
carcinomas using random forest machine learning models (8, 16). We
repeated that analysis but added the concentration of the SCFAs as
possible features to train the models (Figure S1). Models trained using
SCFAs to classify individuals as having adenomas or carcinomas rather
than normal colons had median areas under the receiver operator
characteristic curve (AUROC) that were significantly greater than 0.5
(P\textsubscript{adenoma}\textless{}0.001 and
P\textsubscript{carcinoma}\textless{}0.001). However, the AUROC values
to detect the presence of adenomas or carcinomas were only 0.54 and
0.55, respectively, indicating that SCFAs had poor predictive power on
their own (Figure 2A). When we trained the models with the SCFAs
concentrations and OTU or genus-level relative abundances the AUROC
values were not significantly different from the same models trained
without the SCFA concentrations (P\textgreater{}0.15; Figure 2A). These
data demonstrate that knowledge of the SCFA profile from a subject's
fecal sample did not improve the ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations using a random forest model.} We next asked whether the
fecal community structure was predictive of fecal SCFA concentrations,
regardless of a person's diagnosis. We trained random forest regression
models using 16S rRNA gene sequence data binned into OTUs and genera to
predict the concentration of the SCFAs (Figure S2). The largest
R\textsuperscript{2} between the observed SCFA concentrations and the
modeled concentrations was 0.14, which was observed when using genus
data to predict butyrate concentrations (Figure 2B). We also used a
smaller dataset of shotgun metagenomic sequencing data generated from a
subset of our cohort (N\textsubscript{normal}=27,
N\textsubscript{adenoma}=25, and N\textsubscript{cancer}=26) (18). We
binned genes extracted from the assembled metagenomes into operational
protein families (OPFs) or KEGG categories and trained random forest
regression models using metagenomic sequence data to predict the
concentration of the SCFAs (Figure S2). Similar to the analysis using
16S rRNA gene sequence data, the metagenomic data was not predictive of
SCFA concentration. The largest R\textsuperscript{2} was 0.055, which
was observed when using KEGG data to predict propionate concentrations
(Figure 2B). Because of the limited number of samples that we were able
to generate metagenomic sequence data from, we used our 16S rRNA gene
sequence data to impute metagenomes that were binned into metabolic
pathways or KEGG categories using PICRUSt (Figure S2). SCFA
concentrations could not be predicted based on the imputed metagenomic
data. The largest R\textsuperscript{2} was 0.085, which was observed
when using KEGG data to predict propionate concentrations (Figure 2B).
The inability to model SCFA concentrations from microbiome data
indicates that the knowledge of the abundance of organisms and their
genes was insufficient to predict fecal SCFA concentrations.

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not meaningfully
predictive of SCFA concentrations. These results complement existing
literature that suggest that fiber consumption and the production of
SCFAs are unable to prevent the risk of developing colonic tumors. It is
important to note that our analysis was based on characterizations of
SCFA and microbiome profiles using fecal samples at a single time point.
Furthermore, observations along the mucosa near the site of lesions may
provide a stronger association. This may be a cautionary result to
temper enthusiasm for SCFAs as a biomarker of gut health more generally.
Going forward it is critical to develop additional hypotheses for how
the microbiome and host interact to drive tumorigenesis so that we can
better understand tumorigenesis and identify biomarkers that will allow
early detection of lesions.

@@

\textbf{Random forest models.} The classification models were built to
predict lesion type from microbiome information with or without SCFA
concentrations. The regression models were built to predict the SCFA
concentrations of acetate, butyrate, and propionate from microbiome
information. For classification and regression models, we pre-processed
the features by scaling them to vary between zero and one. Features with
no variance in the training set were removed from both the training and
testing sets. We randomly split the data into training and test sets so
that the training set consisted of 80\% of the full dataset while the
test set was composed of the remaining data. The training set was used
for hyperparameter selection and training the model and the test set was
used for evaluating prediction performance. For each model, the best
performing hyperparameter, mtry, was selected in an internal five-fold
cross-validation of the training set with 100 randomizations. The mtry
parameter represents the number of features randomly sampled from the
available features at a question point in the classification tree
(i.e.~called splits of nodes) that, when answered, lead to the greatest
improvement in classification. Six values of mtry were tested and the
value that provided the largest AUROC or R\textsuperscript{2} was
selected. We trained the random forest model using the selected mtry
value and predicted the held-out test set. The data-split,
hyperparameter selection, training and testing steps were repeated 100
times to get a reliable and robust reading of model prediction
performance. We used AUROC and R\textsuperscript{2} as the prediction
performance metric for classification and regression models,
respectively. We used the randomForest R package (version 4.6-14) as
implemented in the caret R package (version 6.0-81) for developing and
testing our models.

@@

\textbf{Code availability.} The code for all sequence curation and
analysis steps including an Rmarkdown version of this manuscript is
available at \url{https://github.com/SchlossLab/Sze_SCFACRC_mBio_2019/}.

@@

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}
